---
title       : A workflow with
subtitle    : BEFdata + rBEFdata + taxize
author      : Claas-Thido Pfaff
job         : Also showcasing other useful tools (plyr)
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---


```{r setup, cache =F, include = F, eval = T}
opts_chunk$set(cache = F)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
    if (!is.null(n <- options$out.lines)) {
        x = unlist(stringr::str_split(x, "\n"))
        if (length(x) > n) {
            # truncate the output
            x = c(head(x, n), "....\n")
        }
        x = paste(x, collapse = "\n")  # paste first n lines together
    }
    hook_output(x, options)
})

# install.packages("devtools")
require(devtools)
# install_github(username = "befdata", repo = "rbefdata", subdir = "rbefdata")
require(rbefdata)
# devtools::install_github(c("sckott/XMLSchema", "sckott/SSOAP"))
# devtools::install_github("taxize", "ropensci")
# install.packages("taxize")
require(taxize)
require(knitr)
require(plyr)
require(gdata)
# set options
options("width"=90)
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
squish <- function(x) gsub("\\s+", " ", x)
```

## Prerequisites

* Packages

```{r install_rbefdata, eval = F}
install.packages("rbefdata")
```

```{r install_taxize, eval = F}
install.packages("taxize")
```

```{r install_other_packages, eval = F}
install.packages(c("gdata", "plyr"))
```

* String cleaning helpers

```{r define_helpers, eval = F}
# leading/trailing spaces
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
# multiple spaces
squish <- function(x) gsub("\\s+", " ", x)
```

---

## List options (rbefdata)

```{r, cache = F}
require(rbefdata)
bef.options()
```

---

## Set options (rbefdata)

- The URL ([Workshop Portal](http://stage.china.befdata.biow.uni-leipzig.de))
  - Which server to communicate with
  - Here you would set your own URL

```{r eval = T, cache = F}
bef.options(url = "http://stage.china.befdata.biow.uni-leipzig.de")
```

- The credentials ([Your profile](http://stage.china.befdata.biow.uni-leipzig.de/profile))
  - Needed for authentication
  - At least for restricted datasets

```{r eval = T, cache = F}
bef.options(user_credentials = "Pcz3lXF82FDULZP2GG29")
```

---

## Set options (rbefdata)

```{r}
bef.options()
```

--- bg:#EEE

## Your turn

* Fire up R-Studio
* Load the `rbefdata` package

```{r loadrbefdata, eval = F}
require(rbefdata)
```

* Get your [credentials](http://stage.china.befdata.biow.uni-leipzig.de/profile)
  - Check your profile page

* Set the [URL](stage.china.befdata.biow.uni-leipzig.de) to the BEFdata server (`bef.options()`)
  - Check the URL of the BEFdata instance
  - `bef.options(...)`

* Set the Credentials 
  - `bef.options(...)`

---

## Get data

```{r, eval = F}
# the id can be found in the url of the dataset
bef.portal.get.dataset_by(id = xx)
```

* Get dataset from the [portal](http://stage.china.befdata.biow.uni-leipzig.de/datasets/6) with id #6

```{r get_dataset, cache = F}
dataset = bef.portal.get.dataset_by(id = 6)
```

```{r, eval = F}
head(dataset, 4)
```

```{r kable_head_of_dataset, echo = F}
kable(head(dataset, 4))
```

---

## Inspect the data

* Inspect data (metadata)
  
```{r get_dataset_metadata, eval = T, tidy = F}
attributes(dataset)$title
names(attributes(dataset))
```

---

## Inspect data

* Get detailed information
  - On the columns

```{r get_column_description, eval = T, tidy=FALSE}
names(dataset)
attributes(dataset)$column$description
```

--- bg:#EEE

## Your turn

* Get the dataset with the id = 6
  - use `bef.portal.get.dataset_by(id)`
  - Save it to a variable 

* Inspect the dataset metadata
  - find the creator and save it to a variable
  - find the intellectual rights and save it to a variable
  - hint: use `attributes()` and `names()`

--- bg:#EEE

## Your turn (solution)

* Download

```{r, eval = F}
dataset = bef.portal.get.dataset_by(id = 6)
```

* Creator and intellectual rights

```{r}
attributes(dataset)$creators
attributes(dataset)$intellectualRights
```

---

## Aggregate (plyr)

* Load the package

```{r plyr, eval = F}
require(plyr)
```

* Syntax

```{r, eval = F}
ddply(data.frame, variable(s), function, optional arguments)
# summarise vs transform
```

* Clean the column

```{r}
# clean the species column 
dataset$SpeciesFull = trim(dataset$SpeciesFull)
dataset$SpeciesFull = squish(dataset$SpeciesFull)
```

---

## Aggregate (plyr)

* Aggregate by Species full calculate mean c/n ratio

```{r, cache = F}
aggregated_dataset = ddply(dataset, .(SpeciesFull), summarise, 
                           CarbonNitrogenRatio = mean(Carbon/Nitrogen))
```

* Lets have a look

```{r, eval = F}
head(aggregated_dataset, 4)
```

```{r, echo = F}
kable(head(aggregated_dataset, 4))
```

---

## Use taxize (taxonomic tool-belt)

* Taxize offers access to various databases
    - Integrated Taxonomic Information System [itis](http://www.itis.gov/) 
    - National Center for Biotechnology Information [ncbi](http://www.ncbi.nlm.nih.gov/) 
    - International Union for Conservation of Nature [iucn](http://www.iucn.org/)

* But that is not all (e.g create phylogeny tree, see [tutorial](http://ropensci.org/tutorials/taxize_tutorial.html))


* Load ([taxize](http://ropensci.org/tutorials/taxize_tutorial.html))

```{r require_taxize, eval = F, messages = F}
require(taxize)
```

* e.g. retrieve higher taxonomic names, synonyms and many more

---

## Use taxize (taxonomic tool-belt)

* Get family names

```{r, cache = T, messages = F}
ncbi_family_names = tax_name(query = aggregated_dataset$SpeciesFull, 
                             get = "family", 
                             db = "ncbi", 
                             verbose = F)
```

* Lets check that out

```{r, eval = F}
head(ncbi_family_names, 3)
```

```{r, echo = F}
kable(head(ncbi_family_names, 3))
```

---

## Use taxize (taxonomic tool-belt)

* Combine

```{r}
aggregated_dataframe_family = data.frame(aggregated_dataset, 
                                         FamilyNames = ncbi_family_names$family)
```

```{r, echo = F}
kable(head(aggregated_dataframe_family, 2))
```

* Merge

```{r}
merged_dataset_family = merge(dataset, aggregated_dataframe_family[c(1,3)], by = "SpeciesFull")
```

```{r, echo = F}
kable(head(merged_dataset_family, 2))
```

--- bg:#EEE

## Your turn

* Use `unique()` species names to retrieve the family 
  - use the function `tax_name` from the taxize package
  - set the parameters `get="family"`, `db="ncbi"`
* Merge family names back to the downloaded dataframe 
  - use `merge()` function
* Aggregate a mean carbon/nitrogen ratio on family level
  - Use `ddply(dataset, .(ColumnName), function, newColumn = ...)` function 
  - Try out summarise vs transform as function in the ddply call

---

## Solution

```{r solutionCarbonNitrogenOverFamily, cache = T, eval = T, messages = F}
unique_species = unique(dataset$SpeciesFull)
ncbi_family_names = tax_name(query = unique_species, 
                             get = "family", 
                             db = "ncbi", verbose = F)
names(ncbi_family_names) = c("db", "SpeciesFull", "FamilyNames")
merged_dataset_family = merge(dataset, ncbi_family_names[c(2,3)], by = "SpeciesFull")
aggregated_dataset = ddply(merged_dataset_family, .(FamilyNames), summarise, 
                           CarbonNitrogenRatio = mean(Carbon/Nitrogen))
```


```{r, echo = F}
kable(head(aggregated_dataset, 4))
```

---

## Upload dataset (rBEFdata)

* Upload the data to the portal

```{r, eval = F}
# provide a dataset and title
bef.portal.upload.dataset(...)
```

* Upload the aggregated carbon/nitrogen dataset 

```{r upload_corrected, eval = F}
title = "Carbon nitrogen ratio by family (claas)"
bef.portal.upload.dataset(dataset = aggregated_dataset, dataset_title = title)
```

```
Your data has been uploaded successfully!
You can find your dataset now under the id:7
```

--- bg:#EEE

## Your turn

* Upload the new dataset to the BEFdata server
  - Provide an own unique title (Maybe add your name in the title) 
  - use: `bef.portal.upload.dataset(dataset = , dataset_title = )

---

## Thanks four your attention
